---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# votevizr

<!-- badges: start -->
<!-- badges: end -->

The `votevizr` package helps you make diagrams to better understand election results in *ordinal voting systems*, a class that includes 

- [positional methods](https://en.wikipedia.org/wiki/Positional_voting), including
    + plurality
    + Borda count
    + anti-plurality
- [Condorcet methods](https://en.wikipedia.org/wiki/Condorcet_method), with many ways of resolving cycles
- [ranked-choice voting](https://en.wikipedia.org/wiki/Instant-runoff_voting) (aka RCV, instant-runoff voting/IRV, alternative vote/AV, single-winner STV, preferential voting)   

`votevizr` implements the methods I develop in [this paper](http://andy.egge.rs/papers/diagramming_election_results_v7_for_RR2.pdf), which has been accepted at the journal [*Social Choice and Welfare*](https://www.springer.com/journal/355). The paper explains the figures in detail and discusses previous attempts to represent election results in ternary diagrams. 

## Installation

To install from GitHub: 

``` r
devtools::install_github("aeggers/votevizr")
```

## Example

The package includes two datasets, `brexit_prefs` and `sf_result`. Each dataset is just a named list containing frequencies for each preference ordering over three alternatives. 

We load the Brexit poll results:

```{r example_data}
library(votevizr)
data(brexit_prefs)
brexit_prefs
```

Now we make some figures. Let's start with plurality: 
```{r fig_1, out.width="75%"}
brexit_prefs %>% 
  qplot_votevizr(split = " > ", method = "plurality")
```

Each vertex of the triangle represents an election result where one alternative is ranked first by all voters; the closer you get to a vertex, the larger that alternative's share of *first preferences* (i.e. top rankings). The black dot represents the actual result: "Remain" was ranked first most often (on just under 50% of ballots), while "Deal" and "No deal" were each ranked first on just over 25% of ballots.

Each colored region identifies the first-preference results where a given candidate would win. In plurality the outcome depends only on first preferences, so these regions are symmetrical.

If we move to Borda count, the *first-preference win regions* become asymmetrical: 
```{r fig_2, out.width="75%"}
brexit_prefs %>% 
  qplot_votevizr(split = " > ", method = "Borda")
```

This is because (unlike plurality) Borda count gives points for being ranked second, and "Deal" was ranked second on most ballots; given this pattern of lower preferences, "Deal" would win a Borda count election even if it was well below a plurality of first-preference votes. 

Note that the figure assumes the pattern of *second preferences* (lower rankings) is fixed: as we move away from the dot, we see what would have happened if e.g. "Remain" had won more first-preference votes but the proportion of "Remain" voters listing "Deal" second (vs. "No deal") was held fixed. 

With Condorcet methods, the winner is determined by checking each pair of candidates: a candidate who is ranked higher than each other candidate by a majority is the winner. In the Brexit case, we essentially have a tie between "Remain" and "Deal":

```{r fig_3, out.width="75%"}
brexit_prefs %>% 
  qplot_votevizr(split = " > ", method = "Condorcet")
```

Note that there is an area with no winner: this is where a [Condorcet cycle](https://en.wikipedia.org/wiki/Condorcet_paradox) takes place, with "Remain" beating "Deal", "Deal beating "No deal", and "No deal" beating "Remain". `votevizr` allows several ways of resolving such a cycle. Here is the Kemeny-Young method:  

```{r fig_4, out.width="75%"}
brexit_prefs %>% 
  qplot_votevizr(split = " > ", method = "Condorcet", if_cycle = "kemeny")
```

Finally we have ranked-choice voting: 

```{r fig_5, out.width="75%"}
brexit_prefs %>% 
  qplot_votevizr(split = " > ", method = "RCV")
```

## Vignettes

`vignette("overview")` gives a more detailed overview of the package focusing on the `qplot_votevizr()` method. 

`vignette("under_the_hood")` shows what is happening "under the hood", which is useful if you want more plotting flexibilty than `qplot_votevizr()` provides.
