---
title: "`votevizr`: under the hood"
author: "Andy Eggers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{underthehood}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 5,
  fig.align = "center"
)
```

The `votevizr` package provides tools for diagramming ordinal election outcomes given preference data.

The [Overview](overview.html) vignette shows how to make diagrams using the `votevizr::qplot_votevizr()` convenience function. 

This vignette is for "power users" who want more flexibility.   

## Inputting preference data 

We will use the November 2018 Brexit poll result, which we store as a named list: 
```{r pref_data, echo = T}
brexit_result <- list("Remain > Deal > No deal" = .375, 
               "No deal > Deal > Remain" = .228, 
               "Deal > No deal > Remain" = .212, 
               "Remain > No deal > Deal" = .087, 
               "Deal > Remain > No deal" = .059, 
               "No deal > Remain > Deal" = .038)
```


## Extracting vertices of the first-preference win regions 

```{r setup_1, results = "hide", message = F, warning = F}
library(votevizr)
```

The `votevizr::first_preference_win_regions()` function takes the election/poll result as an argument and returns a dataframe with the vertices of each candidate's first-preference win region in a given voting system.

An example:  

```{r brexit_illustration_1, echo = T}
brexit_rcv <- first_preference_win_regions(brexit_result, split = " > ", method = "RCV")

knitr::kable(brexit_rcv)
```

Each row in `brexit_rcv` provides the coordinates for a vertex on the first-preference win region of a candidate (specified in the `candidate` column). The candidate names are extracted from the names of the `brexit_result` list. 

At this point you could make a figure using base `R`.[^baseR]  

[^baseR]: This was my initial approach, so I have a lot of spare to code to share -- get in touch if you're interested.

## Plotting first-preference win regions with `ggplot2`

Because the output of `votevizr::first_preference_win_regions()` is in "tidy" format, we can more easily plot the first-preference win regions in `ggplot2`. 

```{r setup_2, results = "hide", message = F, warning = F}
library(ggplot2)
library(dplyr)
```

Here is the most basic version: 

```{r plot_brexit_basic}
brexit_rcv %>% 
  ggplot(aes(x = Remain, y = Deal)) +
  geom_polygon(aes(group = candidate, fill = candidate)) + 
  labs(fill = "Winning\ncandidate")
```

I prefer to plot the results in barycentric or ternary coordinates, because it treats the three alternatives more symmetrically. 

The `ggtern` package is specifically for making ternarty diagrams in `R`. But `ggtern` causes conflicts with `ggplot2` and has lots of functionality we don't need, and it is surprisingly easy to make a ternary diagram without it. So we'll proceed with `ggplot2` alone.  

To convert data from the simplex in $(x, y, z)$ coordinates into ternary coordinates $(x', y')$, the transformation is
\begin{eqnarray}
x' &:=& x + \frac{1}{2}y \\
y' &:=& \frac{\sqrt{3}}{2} y. \\
\end{eqnarray} 

So first do that transformation:
```{r brexit_ternary_transform}
brexit_rcv %>% 
  mutate(Remain = Remain + .5*Deal, Deal = sqrt(3/4)*Deal) ->
  brexit_rcv_ternary
```

Then plot it, with some added style:
```{r plot_brexit_ternary}
brexit_rcv_ternary %>%
  ggplot(aes(x = Remain, y = Deal)) +
  geom_polygon(aes(group = candidate, fill = candidate), show.legend = F) +
  coord_fixed() + # ensures we get an equilateral triangle
  scale_fill_brewer() + 
  theme_void() + 
  expand_limits(x = c(-.1, 1.1), y = sqrt(3/4)*c(-.1, 1.1)) + 
  annotate(geom = "text", x = c(1,.5,0), y = c(0,sqrt(3/4),0) + .05*c(-1,1,-1), label = c("Remain", "Deal", "No deal")) -> p
p
```

## Convenience functions for plotting

We can also add a dot for the observed result, using the `votevizr::first_preference_shares()` method to extract the observed first-preference shares: 
```{r brexit_with_result}
brexit_fps <- brexit_result %>% 
  first_preference_shares(split = " > ") %>% 
  mutate(Remain = Remain + .5*Deal, Deal = sqrt(3/4)*Deal)

p2 <- p + geom_point(data = brexit_fps, aes(x = Remain, y = Deal), cex = .75)
p2
```

Now add gridlines using `votevizr::geom_ternary_gridlines()`: 
```{r add_gridlines}
p2 + 
  geom_ternary_gridlines()
```

## Handling Condorcet completion methods 

`votevizr` has two ways of handling Condorcet completion methods, i.e. rules for determining a winner in the event of a Condorcet cycle.

The user specifies the completion method (both for `first_preference_win_regions()` or `qplot_votevizr()`) with the `if_cycle` argument.

- when `if_cycle` is `"empty"`, `"kemeny"`, or the name of a candidate, then `first_preference_win_regions()` specifies the complete first-preference win regions and `qplot_votevizr()` simply plots them. 
- when `if_cycle` is `"RCV"`, `"IRV"`, `"plurality"`, `"borda"`, `"anti-plurality"`, or `"positional"`, then `first_preference_win_regions()` leaves the cyclic triangle empty; `qplot_votevizr()` plots the first-preference win regions for the *completion method* (e.g. RCV when `if_cycle = "RCV"`) first and then overlays the Condorcet first-preference win regions.[^why]

[^why]: There may be a clever way to make `first_preference_win_regions()` get the first-preference win regions for both Condorcet and the completion method and combine them correctly, but for plotting it was easier to just overlay two plots. 

Let's illustrate this second approach for the case where the Brexit winner is to be determined by Condorcet with RCV used in the event of a cycle. 

First let's show the diagram with the Condorcet cycle not filled in. Get the first-preference win regions and transform them for the ternary diagram:

```{r brexit_condorcet}
brexit_result %>% 
  first_preference_win_regions(split = " > ", 
                               method = "condorcet") %>% 
  mutate(Remain = Remain + .5*Deal, Deal = sqrt(3/4)*Deal) -> brexit_condorcet 
```

Now, reusing code from above, we make the plot:
```{r brexit_condorcet_plot}
brexit_condorcet %>% 
  ggplot(aes(x = Remain, y = Deal)) +
  geom_polygon(aes(group = candidate, fill = candidate), show.legend = F) +
  coord_fixed() +
  scale_fill_brewer() + 
  theme_void() + 
  expand_limits(x = c(-.1, 1.1), y = sqrt(3/4)*c(-.1, 1.1)) + 
  annotate(geom = "text", x = c(1,.5,0), y = c(0,sqrt(3/4),0) + .05*c(-1,1,-1), label = c("Remain", "Deal", "No deal")) 
```

Note the empty triangle. For first-preference results in this area (holding fixed the pattern of lower preferences), there is a Condorcet cycle: "Remain" beats "Deal", "Deal" beats "No deal", and "No deal" beats remain. 

Now suppose the winner is chosen by RCV if there is a cycle. We illustrate this rule by redrawing the RCV plot from above (stored as `p`) and then overlaying the Condorcet first-preference win regions: 

```{r brexit_condorcet_rcv}
p +   # p was the RCV figure 
  geom_polygon(data = brexit_condorcet, aes(group = candidate, fill = candidate), show.legend = F)  # overlay the Condorcet figure
```

We can do the same using any positional method (Borda, plurality, antiplurality, or others). 

To produce the same figure via `qplot_votevizr()`: 
```{r brexit_condorcet_rcv_2}
brexit_result %>% 
  qplot_votevizr(split = " > ", method = "condorcet", if_cycle = "RCV")  
```

## Want to know more?

Read my paper, forthcoming in *Social Choice and Welfare* (preprint [here](http://andy.egge.rs/papers/diagramming_election_results_v7_for_RR2.pdf)), which explains the figures in more detail and discusses previous attempts to represent election results in ternary diagrams. 

See also the [Overview](overview.html) vignette for some basic background about the figures and more on the `qplot_votevizr()` function. 

